---
# Change the title etc. to your needs:
title: "Bayesian Rating Models"
subtitle: "Starcraft II"
author: "Adam Frederik Ingwersen Linnemann, 2018952670"
course: 'Introduction to Bayesian Statistics, Korea University'
logo: ku-logo.png  # insert path to your logo


abstract: |
  This report investigates the power of bayesian rating models in predicting professional Starcraft II league matches. Prediction on outcomes for the 2017 Korean championships are being considered; a variety of distributional relations are proposed, but it is found that the modelling of attacking and defensive parameters are suitable - and should follow a normal distribution. The proposed model rightly predicts the winner of the 2017 championships - but only some of the runner-ups are predicted correctly. More complicated models are proposed to alleviate this drawback of the base model.
  
# Change the following lines only if you know what you are doing:
date: '`r format(Sys.Date(), "%d\\. %m\\. %Y")`'  # today
fontfamily: lmodern
fontsize: 11pt
graphics: null
papersize: 
geometry: margin=1.0in
classoption:
  - a4paper
  - oneside
  #- more options here, see rmarkdown documentation 
lang: en-US
toc: yes
numbersections: yes
UP_title: yes
UP_subtitle: yes
shaded_quote: no
output: 
  yart::yart
---

```{r setup, include=FALSE, echo = FALSE, warning = FALSE, fig.align="center"}
knitr::opts_chunk$set(echo = TRUE, fig.pos= "h")
library(knitr)
library(kableExtra)
options(kableExtra.latex.load_packages = T)
```

# Scope and Purpose
The aim of this report is to investigate the performance and interpretability of bayesian rating models, as proposed by Baio & Blangiardo (2009), on Starcraft II match data. 

# Data
Data on Starcraft II matches from 2014 to 2017 is obtained from Kaggle. The data contains 187.397 match results played between 1.856 distinct players (professional and non-professional). 

\begin{figure}
\centering
\includegraphics[width=1.0\linewidth]{img/table_1} \caption{Raw data}\label{fig:table1}
\end{figure}

In order to achieve a sensible amount of data for MCMC simulation on a laptop, the data has to be subsetted. This is done by looking at match-dates after \textit{2016-01-01}. 
Furthermore, the number of players are limited to 8, by grabbing the top 8 players in the \textbf{2017 Starcraft II World Championship} - where the top 8 constitutes the finalists of that tournament. 

```{python, eval = F}
xpath = "//tr/td[@align = 'left']/a/text()"
url = "https://liquipedia.net/starcraft2/2017_StarCraft_II_World_Championship_Series_Korea/Standings"
page = requests.get(url)
tree = html.fromstring(page.content)
top  = tree.xpath(xpath)
top8 = top[:8]
``` 

Further data preprocessing steps are documented in the attached \texttt{report.ipynb}. 
\newline
The players' defensive and offensive performance is represented in the figure below for the two considered seasons.
\begin{figure}
\centering
\includegraphics[width=1.1\linewidth]{img/subplot_2} \caption{Relative player def/off performace}\label{fig:subplot2}
\end{figure}

In addition, the proportionality should ideally be somewhat equivalent between players. It's apparent that \textbf{Rogue} and \textbf{TY} are slightly underrepresented in the data. 

```{r, eval = T, echo = F}
players = cbind(c('Stats', 'INnoVation', 'Dark', 'soO', 'TY', 'GuMiho', 'herO', 'Rogue'))
proportion = cbind(c(0.14977307110438728,
 0.15279878971255673,
 0.11800302571860817,
 0.12405446293494705,
 0.08169440242057488,
 0.15431164901664146,
 0.15128593040847202,
 0.0680786686838124))
dt = data.frame(players, proportion)
knitr::kable(dt, caption ="Player proportion of total dataset") %>%
  kable_styling(latex_options = "h")
```

# Modelling

In attempting to model the relations present in a Starcraft II tournament and capture the inhrerent dynamics that apply to such competitions, three distinct models are proposed. Each of which are to be explained in sections below.
All models rely on having a set of proxy-distributions that represent the offensive and defensive attributes of any particular player. The model statistics and related plots are presented in the attached \texttt{ipython-notebook}.

## Baseline Model

This model serves as the basline to which other models are evaluated. It's a stripped down version of the model presented by \textit{Baio \& Blangiardo (2010)}, that takes a fixed zero-mean in offensive/defensive attributes, and only the parameter variance is modeled explicitly. 

### Specification

```{python, eval = F}
tau_att    = pm.Gamma('tau_att', 0.1, 0.1)
tau_def    = pm.Gamma('tau_def', 0.1, 0.1) 
atts_star = pm.Normal("atts_star", mu = 0, sd = tau_att, shape = n_players)
defs_star = pm.Normal("defs_star", mu = 0, sd = tau_def, shape = n_players) 
atts = pm.Deterministic('atts', atts_star - tt.mean(atts_star))
defs = pm.Deterministic('defs', defs_star - tt.mean(defs_star))
p1_theta = tt.exp(atts[player_1] + defs[player_2])
p2_theta = tt.exp(atts[player_2] + defs[player_1])
p1_points = pm.Poisson('p1_points', mu = p1_theta, observed = observed_player1_score)
p2_points = pm.Poisson('p2_points', mu = p2_theta, observed = observed_player2_score)
``` 

### Results

With NUTS-sampling, the model converges quickly and achieves excellent \textbf{Rubin-Gelman} statistics and \textbf{BFMI (Bayesian Fraction of Missing Information)}, which indicates that the model both converged and is well-specified, respectively. 

## Intercept Model

This model is a re-write of the model presented in \textit{Baio \& Blangiardo (2010)}, with a tunable intercept to indicate a home-court advantage to one of the teams. This intuitively should not provide any additional information in a Starcraft II context.

### Specification

```{python, eval = F}
tau_att    = pm.Gamma('tau_att', 0.1, 0.1)
tau_def    = pm.Gamma('tau_def', 0.1, 0.1) 
intercept  = pm.Flat('intercept')
atts_star = pm.Normal("atts_star", mu = 0, sd = tau_att, shape = n_players)
defs_star = pm.Normal("defs_star", mu = 0, sd = tau_def, shape = n_players)
atts = pm.Deterministic('atts', atts_star - tt.mean(atts_star))
defs = pm.Deterministic('defs', defs_star - tt.mean(defs_star))
p1_theta = tt.exp(intercept + atts[player_1] + defs[player_2])
p2_theta = tt.exp(intercept + atts[player_2] + defs[player_1])
p1_points = pm.Poisson('p1_points', mu = p1_theta, observed = observed_player1_score)
p2_points = pm.Poisson('p2_points', mu = p2_theta, observed = observed_player2_score)
```

### Results

This specification has been run several times under the exact same conditions as the baseline model, but does not appear to converge - and may be misspecified as well. This suggests, that the inclusion of the intercept as a trainable parameter only makes MCMC more complicated, and does not provide any additional information. Thus, the homecourt-advantage is not present in the observed Starcraft II mathces, and perheaps in Starcraft II in general. 

## Tunable Mean Model wo. Intercept

This model is, again, inspired by the one suggested by \textit{Baio \& Blangiardo (2010)}, but disregards the intercept, and uses a trainable mean for both the offensive and defensive stats of the players. This seems to be a more sensible approach in the Starcraft II context.

### Specification

```{python, eval = F}
mu_att     = pm.Normal('mu_att', 0)
mu_def     = pm.Normal('mu_def', 0)
tau_att    = pm.Gamma('tau_att', 0.1, 0.1)
tau_def    = pm.Gamma('tau_def', 0.1, 0.1)
atts_star = pm.Normal("atts_star", mu = mu_att, sd = tau_att, shape = n_players)
defs_star = pm.Normal("defs_star", mu = mu_def, sd = tau_def, shape = n_players)
atts = pm.Deterministic('atts', atts_star - tt.mean(atts_star))
defs = pm.Deterministic('defs', defs_star - tt.mean(defs_star))
p1_theta = tt.exp(atts[player_1] + defs[player_2])
p2_theta = tt.exp(atts[player_2] + defs[player_1])
p1_points = pm.Poisson('p1_points', mu = p1_theta, observed = observed_player1_score)
p2_points = pm.Poisson('p2_points', mu = p2_theta, observed = observed_player2_score)
```

### Results

This model performs rather similar to the baseline model, but does yield a lot of divergences in the sampling process, and takes longer to converge - thus it should be discarded in favor of the baseline model.

# Discussion of major findings

The first major finding has been to keep the model simple, as there does not appear to be complicated interactions in the data, that the model has a hard time capturing. Had the data been more detailed or more scarce, it may have been that the baseline model would not suffice. 

## Picking the best model

The best results are, surprisingly, obtained by the baseline model, which shows telltale signs of convergence and proper specification. 

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\linewidth]{img/rubin-gelman} \caption{Specification and convergence of baseline-model}\label{fig:rg1}
\end{figure}

## Starcraft v. Soccer
Obviously Starcraft II is a very different sport from soccer. The fact, that soccer is played in teams, whilst Starcraft II is played by individuals at the competitive level is a good example of the differences pertaining to the sports. As elaborated above, the does not appear to be any homecourt advantage, which is a rather significant factor in determining match outcomes. The proposed modelling framework assumes, that the offensive and defensive properties are seperable in Starcraft II, which may not be the case in the same way as in Soccer. It may very well be the case, that one should include the player activity such as APM (Actions Per Minute) as a factor. In Starcraft II defensive strategies also often rely of offensive harrasment of the opponent, which the proposed modelling framework may have trouble picking up on. 

With that being said, it is not unreasonable to think that there is a tradeoff for players attempting offensive strategies by eroding defensive capabilites in the short term. 

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\linewidth]{img/player-defense} \caption{Player Defensive Capabilites}\label{fig:def1}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\linewidth]{img/player-offense} \caption{Player Offensive Capabilites}\label{fig:off1}
\end{figure}

The defensive capabilites are directly linked to the opponents score - so the lower defensive capabilites, the better the player is at not letting the opponent get points.

# Implications 

The work that has been conducted this far can be trivially used to resample from the model and thus predict the final ranking of the players in the tournament. 

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\linewidth]{img/final-table} \caption{2017 Tournament Predictions}\label{fig:tabl}
\end{figure}

We can compare to the actual final ranking:

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\linewidth]{img/final-ranking} \caption{2017 Tournament Results}\label{fig:tabl2}
\end{figure}

# Concluding Remarks

This project has shown that bayesian rating models certainly are applicable to predicting Starcraft II rankings, although the methodology applied falls a bit short. 

Future projects should seek to explore binomial distributions for the offensive and defensive capabilites and focus on a smaller player pool. 


<!-- These lines ensure references are set with hanging indents in PDF documents -->
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}



